---
title: "Assignment 1"
format: html
execute:
  cache: false 
---

# Task 1: Tie Dependence and CUGs

## Setup

```{r setup}
library(sna)
library(network)

# Given by task statement
set.seed(161)
n_perm <- 5000

n <- 22
m <- 134
M <- 42
A <- 50
N <- 139

# Data path from this folder to the project Data folder
data_path <- "1. Tie Dependence and CUGs/Lintner_Data/10_W1.csv"
```

#### Load Network and Remove Affected Data

```{r load-network}
# 1) Load network with blanks as NA
raw_net <- read.csv2(
  data_path,   # adjust path if needed
  header = TRUE,
  row.names = 1,
  check.names = FALSE,
  na.strings = c("", "NA")
)

# Keep original node labels and normalize possible "X" prefixes in numeric IDs
row_ids <- rownames(raw_net)
col_ids <- colnames(raw_net)

# Convert to numeric matrix and align to common IDs on both axes
A_raw <- as.matrix(raw_net)
storage.mode(A_raw) <- "numeric"
rownames(A_raw) <- row_ids
colnames(A_raw) <- col_ids

common_ids <- intersect(row_ids, col_ids)
A_raw <- A_raw[common_ids, common_ids, drop = FALSE]

# 2) Identify nodes affected by missing tie data
nodes_drop <- rownames(A_raw)[rowSums(is.na(A_raw)) > 0]

# 3) Remove affected nodes from network (row + column)
A_clean <- A_raw[!(rownames(A_raw) %in% nodes_drop), !(colnames(A_raw) %in% nodes_drop), drop = FALSE]

# 4) Final checks
stopifnot(nrow(A_clean) == ncol(A_clean))
stopifnot(all(A_clean %in% c(0, 1)))
```

#### Summary Table

```{r summary-table}
summary_table <- data.frame(
  metric = c("Original nodes", "Dropped nodes", "Cleaned nodes", "Remaining NA", "Ties in cleaned network"),
  value = c(
    nrow(A_raw),
    length(nodes_drop),
    nrow(A_clean),
    sum(is.na(A_clean)),
    sum(A_clean)
  )
)

nodes_drop_table <- data.frame(
  dropped_node = if (length(nodes_drop) == 0) NA_character_ else nodes_drop,
  stringsAsFactors = FALSE
)

knitr::kable(summary_table, caption = "Cleaning Summary")
knitr::kable(nodes_drop_table, caption = "Dropped Node IDs")
```

#### Cleaned 10_W1 CSV

```{r save-clean-csv}
# Save cleaned adjacency matrix as a CSV next to this qmd
clean_csv_path <- "10_W1_clean.csv"
write.csv2(
  data.frame(node = rownames(A_clean), A_clean, check.names = FALSE),
  file = clean_csv_path,
  row.names = FALSE,
  quote = FALSE
)
```

## Task 1.1

*Compute probability* $p$ of observing a tie if you choose one possible dyad at random

```{r task-1-1-tie-prob}
nodes = n
ties  = m

# P[tie] = |ties| / |total number of possible ties|
n_dyads = nodes * (nodes - 1) / 2
n_ties = nodes * (nodes - 1)
p = ties / n_ties
cat("Probability[Observing a Tie] = ", p, "\n")
```

## Task 1.2

*Compute the probabilities* $p_M, p_A, p_N$ of observing a mutual, asymmetric and null dyad

```{r task-1-2-dyad-probs}
n_mutual     <- M
n_asymmetric <- A
n_null       <- N

p_M <- n_mutual     / n_dyads
p_A <- n_asymmetric / n_dyads
p_N <- n_null       / n_dyads

knitr::kable(
  data.frame(
    Dyad        = c("Mutual", "Asymmetric", "Null", "Total"),
    Count       = c(n_mutual, n_asymmetric, n_null, n_dyads),
    Probability = c(p_M, p_A, p_N, p_M + p_A + p_N)
  ),
  digits = 4,
  caption = "Dyad Census"
)
```

## Task 1.3

*What would be the value of the probabilities in (2) if we assume tie independence and that the probability of observing a tie takes the value* $p$ computed in (1)?

```{r task-1-3-tie-prob-uniform}
# T : Tie
# N : No Tie
#
# p[T] = p
# p[N] = 1 - p
#
#      .
#     / \
#   T     N
#  /\     /\
# t  N   T  N

p_uni_M <- p * p
p_uni_A <- 2 * p * (1 - p)
p_uni_N <- (1 - p) * (1 - p)

knitr::kable(
  data.frame(
    Dyad        = c("Mutual", "Asymmetric", "Null"),
    Probability = c(p_uni_M, p_uni_A, p_uni_N)
  ),
  digits = 4,
  caption = "Dyad Census"
)
```

## Task 1.4

*Given the values obtained in (2) and (3), would it be reasonable to assume tie independence when performing further analysis on the observed network? Justify*

No, it would not be reasonable to assume tie independence.

```{r task-1-4-comparison}
observed    <- c(p_M, p_A, p_N)
independent <- c(p_uni_M, p_uni_A, p_uni_N)

knitr::kable(
  data.frame(
    Dyad        = c("Mutual", "Asymmetric", "Null"),
    Observed    = observed,
    Independent = independent,
    Factor      = observed / independent
  ),
  digits = 4,
  caption = "Observed vs. Independence Assumption"
)
```

As shown in the table we observe roughly twice as many mutual ties as we would by assuming independence. On the other hand asymmetric ties would be overrepresented by the assumption of independence.

## Task 1.5

*Consider a different model that assumes all nodes have an outdegree of nine (d=9).* *Compute the probability of a mutual tie assuming tie independence. Compare this probability with the probabilities computed in (2) and (3). Would it be reasonable to assume this model?*

```{r task-1-5-different-model}
m_nine <- n * 9
p_nine <- m_nine / n_ties
cat("Probability[Observing a Tie given (d=9)] = ", p_nine, "\n")

p_nine_M <- p_nine * p_nine
p_nine_A <- 2 * p_nine * (1 - p_nine)
p_nine_N <- (1 - p_nine) * (1 - p_nine)

knitr::kable(
  data.frame(
    Dyad        = c("Mutual", "Asymmetric", "Null"),
    Observed    = c(p_M, p_A, p_N),
    `d=9`       = c(p_nine_M, p_nine_A, p_nine_N),
    Factor      = c(p_M, p_A, p_N) / c(p_nine_M, p_nine_A, p_nine_N)
  ),
  check.names = FALSE,
  digits = 4,
  caption = "Observed vs. d=9 Independence Model"
)
```

**Comment about this model:**\
As we can see this model aligns pretty well the probability of observing a mutual dyad. On the other hand are asymmetric and null dyads not closely aligned.

If we think about how many ties there are: $22 \cdot 9 = 198$, we see it is a lot more than the actual number of ties which is $m = 134$.\
Logically (by assuming tie independence) we would need to pick $d = m/n = 134/22 \approx 6$.\

*Conclusion*: We can match one statistic (in that case mutuality) by tuning a parameter, but the overall dyad distribution still doesn't fit. This means this model isn't adequate either.

## Task 1.6

*Load the packages `sna` and `network` in R. Import the file `10_W1.csv` as a matrix object. The R function `cug.test` perform a CUG test in R. Run the following commands in R:*

```{r task-1-6-prelimimary-loading}
cguRec <- cug.test(A_clean, grecip, cmode = "edges", reps=n_perm)
cugInd <- cug.test(A_clean, centralization, cmode="edges", FUN.arg=list(FUN=degree, cmode="indegree"), reps=n_perm)
cugTrans <- cug.test(A_clean, gtrans, cmode = "dyad.census", reps=n_perm)
```

### Task 1.6.1

*State the hypotheses and the conditional features of the tests* (*Use the help function `?cug.test` to understand the command lines*)

All three tests share the same structure: we compare an observed network statistic against a distribution of that statistic computed over random graphs that preserve a specific feature of the observed network.

The first test (`cguRec`) measures reciprocity (`grecip`) conditioned on the number of edges (`cmode = "edges"`). Random graphs are generated with the same edge count as the observed network.\
$H_0$: The observed reciprocity is no more extreme than expected in random networks with the same number of edges.\
$H_A$: The observed reciprocity is more extreme than expected.

The second test (`cugInd`) measures indegree centralization (`centralization` with `FUN = degree, cmode = "indegree"`), also conditioned on the number of edges.\
$H_0$: The observed indegree centralization is no more extreme than expected in random networks with the same number of edges.\
$H_A$: The observed indegree centralization is more extreme than expected.

The third test (`cugTrans`) measures transitivity (`gtrans`) conditioned on the dyad census (`cmode = "dyad.census"`). Random graphs are generated preserving the exact counts $M$, $A$, $N$.\
$H_0$: The observed transitivity is no more extreme than expected in random networks with the same dyad census.\
$H_A$: The observed transitivity is more extreme than expected.

### Task 1.6.2

*Interpret the results*

```{r task-1-6-2-results}
alpha <- 0.05

cug_summary <- data.frame(
  Test        = c("Reciprocity", "Indegree Centralization", "Transitivity"),
  Conditioning = c("edges", "edges", "dyad.census"),
  `Observed Value` = c(cguRec$obs.stat, cugInd$obs.stat, cugTrans$obs.stat),
  `Pr(X>=Obs)`     = c(cguRec$pgteobs, cugInd$pgteobs, cugTrans$pgteobs),
  `Pr(X<=Obs)`     = c(cguRec$plteobs, cugInd$plteobs, cugTrans$plteobs),
  Decision    = ifelse(
    pmin(c(cguRec$pgteobs, cugInd$pgteobs, cugTrans$pgteobs),
         c(cguRec$plteobs, cugInd$plteobs, cugTrans$plteobs)) < alpha,
    "Reject H0", "Fail to reject H0"
  ),
  check.names = FALSE
)

knitr::kable(cug_summary, digits = 4, caption = "CUG Test Results (alpha = 0.05)")
```

```{r task-1-6-2-plots, fig.width=10, fig.height=4}
par(mfrow = c(1, 3))
plot(cguRec, main = "CUG: Reciprocity\n(conditioned on edges)")
plot(cugInd, main = "CUG: Indegree Centralization\n(conditioned on edges)")
plot(cugTrans, main = "CUG: Transitivity\n(conditioned on dyad census)")
```

As shown in the table, the reciprocity test yields $Pr(X \geq Obs) \approx 0$, so we **reject** $H_0$. The observed reciprocity (0.77) is significantly higher than what random networks with the same density produce. This is consistent with our findings in Tasks 1.2–1.4 where mutual ties were roughly twice as frequent as expected under independence.

For indegree centralization we get $Pr(X \geq Obs) \approx 0.77$, so we **fail to reject** $H_0$. The observed centralization (0.20) falls well within the range of random networks with the same edge count. The distribution of popularity across students is not significantly more unequal than what we would expect by chance.

The transitivity test yields $Pr(X \geq Obs) \approx 0$, so we **reject** $H_0$. Even after controlling for the dyad census, the observed transitivity (0.54) is significantly higher than expected. This means the network exhibits "friend of a friend" clustering beyond what reciprocity alone would produce — if $i \to j$ and $j \to k$, then $i \to k$ is more likely than expected.

# Task 2: MR-QAP Regression

## Code used to solve the tasks

```{r}
#| include: false
source("2. MR-QAP regression/01_DataPrep.R")
source("2. MR-QAP regression/02_Covariates.R")
source("2. MR-QAP regression/03_MRQAP.R")
```

## 2.1 Classroom 10 Basic QAP

After cleaning the data and formatting it in a usable way this is the code that returns the result

```{r}
print(summary(nlTask1))
```

If one prints out the summary, we get that the friendship nominations in wave 2 are highly associated with friendship nominations in wave 1. The chances of a friendship persisting to the second wave if it was already established by the time of the first wave as compared to it not existing in the first wave is **roughly 15 fold**. Having a \*\*\*p-value (Pr(\>=b)) below 0.001 makes this result **highly significant**.

```         
Coefficients:
            Estimate   Exp(b)     Pr(<=b) Pr(>=b) Pr(>=|b|)
(intercept) -0.1663907  0.8467153 0.3778  0.6222  0.3814   
x1           2.7398508 15.4846743 1.0000  0.0000  0.0000   
```

## 2.2 Co-variate Model Specification

Variables x2-x4 represent the following:

**x2: literacyRec**: `matrix(attrFinal[, "literacy_end"], nNodes, nNodes, byrow=FALSE)`\
With the hypothesis, that high literacy students (column j) should get fewer nominations, so a negative effect.

**x3: sameGender**: `outer(attrFinal[, "gender"], attrFinal[, "gender"], "==") * 1`\
With the hypothesis, that same gender pairs (i,j) are more likely friends, so an expect positive effect.

**x4: hiseiSend**: `matrix(attrFinal[, "HISEI"], nNodes, nNodes, byrow=TRUE)`\
With the hypothesis, that High HISEI senders (row i) are to send more nominations, so an expect positive effect.

Comparing the different waves and the attribute table and removing any student that does not have complete data in all of then leaves us with the cleaned and relevant data for the MR-QAP. With that, one can build the matrices.

## 2.3+2.4 Full Model Results

This code outputs the results for all the 8 classrooms for which data was provided

```{r}
print(allResults)
```

For the sub-task 2.3 one can look at the 10.x rows. As seen before, the significance of friendships from wave 1 carrying over to wave 2 remains very high. We can see a small drop off in the beta from \~2.74 to \~2.4 which can be due to the controlling for literacy/gender/HISEI. As for the hypothesis:

-   The hypothesis that students with high ending literacy scores are less likely to receive friendship nominations **can not be supported**. While we can see a negligibly small multiplier, the p-value of 0.8646 makes this **statistically insignificant**.

-   The hypothesis that a friendship nomination is more likely between a pair or students of the same gender **can be supported**. For students with the same gender, the probability of forming a friendship (same gender dyads), has **almost 7x higher odds**. This result is **statistically highly significant** as indicated by the \*\*\*p-value\<0.001.

-   The hypothesis, that students with high HISEI send more friendship nominations **can be supported** for the 10th class. We have roughly **4.7% increase** of sending nomination to any receiver for each 1 unit increase in the sender's HISEI. This result is **statistically significant** with a \*\*p-value\<0.01.

## 2.5 Cross-Classroom Patterns

Analyzing the numbers for all the rooms results in the following conclusions and overarching patterns:

-   Firstly, the stability of friendships seems to remain universally strong and very significant. With 7/8 \*\*\*p-values\<0.001, last one \~0.0016. The odds ratios range from a low of \~3.5x to a high of over 23x. The takeaway is that friendships persist regardless of class characteristics.

-   Secondly, the gender homophily, so same-gendered friendships, is pretty robust. 4/8 classes show a very high significance of \*\*\*p-value\<0.001 while one falls in the category of \*\*p-value\<0.01 and one in the category of \*p-value\<0.05. Only two classes can be considered exceptions, class 4 and class 12. Both have a p-value far greater then 0.1, making them statistically insignificant. As for the significant ones, they range from a low of \~2.2x to a high of \~7.75x times likelihood to form a friendship when students have the same gender.

-   Thirdly, if we do not only look at the 10th class, but rather at all 8 for which we have data, we see a slight change in the conclusion one might draw regarding the HISEI hypothesis. Only class 1, 9, 10 and 12 show any significance in the results. They range from a \*\*p-value of 0.0018 up to a \*p-value of 0.039. Thus, having half the classes give insignificant results is noteworthy and might change ones takeaway from the exception that is class 10. Furthermore, we can see that the multiplier value between the four significant classes points us in both directions, two show an increase, the other two a decrease in the likelihood of sending friendship nominations, ranging from \~0.96x to \~1.046x. All in all, the result we got from only looking at class 10 regarding the hypothesis is questionable at best. Generalizing it would be a mistake.

-   The final variable we looked at was the ending literacy scores. There is no class here that showed any statistical significance, giving us no support for the hypothesis. And even if there was any, the effect of the literacy scores is negligible, never changing the outcome more then 2%. Thus we can say that literacy does not predict popularity.

As for the generalizability: The w1 stability and same Gender homophily seem to be highly generalizable, matching results from studies done on this topic. The literacy is universally irrelevant. HISEI is context-specific and results in mixed directions, so one should take it with a grain of salt.


# Idea

The Network Autocorrelation Model assumes the data is distributed like:
$$y = X\beta + \rho W y + \epsilon,$$
where in our case $y$ is the ending literacy score, $X$ is the other attributes, $W$ is the wave 2 adjacency matrix and $\epsilon$ is distributed normally.

So we just need to statistically estimate $\beta, \rho$.

# Importing and cleaning Data

We need to clean the data by removing all students with missing data. That can be missing friendship data, missing HISEI, missing gender, missing literacy, ...

The important code-block for this cleaning is the following. We get all student ids and intersect them to get only the ones with complete data.

```{r}
#| eval: false
ok_ids1 <- rownames(w2net)[complete.cases(w2net)]
ok_ids2 <- rownames(attr)[complete.cases(attr)]

ok_ids <- intersect(ok_ids1, ok_ids2)
```

The remaining code of `import_and_clean.R` will be uploaded along the pdf.

```{r}
#| message: false
source("3. Network Autocorrelation Model/import_and_clean.R")
w2netClean<-getNet("10")
attrClean<-getAttr("10")
```

# Subtask 1

Subtask 1 requires just Network Autocorrelation, without exogenous variables, so the model simplifies to
$$y = \mathbf{1} \beta + \rho W y + \epsilon.$$
for some constants $\beta, \rho$ to be estimated.

So we fit a NAM model:
```{r}
X <- model.matrix(literacy_end~1, data=attrClean)
model <- lnam(y=attrClean$literacy_end, x=X, W1=w2netClean)
print(summary(model))
```

# Subtask 2

Subtask 2 adds HISEI and gender, giving the model described in **Idea**.

Gender is a simple binary variable, encoding it as $0$ for male and $1$ for female should be sufficient:

```{r}
attrClean["gender"] <- lapply(attrClean["gender"], \(x) x=="female")
```

Now we can basically reuse the code from **Subtask 1**, just adding gender, HISEI and a constant factor for the intersect in the $X$ matrix.

```{r}
X <- model.matrix(literacy_end~ gender + HISEI, data = attrClean)
model <- lnam(attrClean$literacy_end, X, w2netClean)
print(summary(model))
```

# Subtask 3

For classroom 10, we get a negative autocorrelation term with $p$-value $0.028$. So it's significant. This doesn't support the positive autocorrelation hypothesis, it actually contradicts it. This suggests that students are likely connected with other students which have differing literacy scores. Further, being female and having a high HISEI score both correlate positively with the literacy_end score.

<!-- The outcome actually suggests that the autocorrelation is negative. Further, female students outperform male students, and there is a strong correlation between the HISEI score and the literacy_end score. -->

# Subtask 4

**TODO (discuss with others)**

QAP and NAM address different questions. The QAP task examines how the number of friendship nominations correlates with the other variables, while NAM addresses the question wether the ending literacy score correlates with the other variables (especially the network structure itself: "Do high literacy people have high literacy friends?").

So comparing the results of the QAP task with the NAM task does not really make sense.

<!-- Comparing these results with the results from QAP, we see that QAP also revealed that female students perform better, and that there is a correlation between HISEI and literacy_end scores. -->

# Subtask 5

Combining all the above code and running it in a loop over all classrooms, we get:

```{r}
for (classroom in c("01", "02", "03", "04", "05", "09", "10", "12")) {
    w2netClean<-getNet(classroom)
    attrClean<-getAttr(classroom)

    attrClean["gender"] <- lapply(attrClean["gender"], \(x) x=="female")    

    X <- model.matrix(literacy_end~ gender + HISEI, data = attrClean)
    model <- lnam(attrClean$literacy_end, X, w2netClean)
    cat(paste("Classroom number:", classroom, sep=" "))
    cat("\n")
    print(summary(model))
    cat("============================================\n\n\n")
}
```

# Subtask 6

In the results from subtask 5 most of the autocorrelation terms are negative, but all with very large $p$-value. So the result is not statistically significant. Only classroom 10 has evidence for such a negative autocorrelation. These results are not generalizable. So there is no evidence to support a positive autocorrelation hypothesis.

The HISEI and gender correlations also don't generalize to the other classroom. With the exception of classroom 1, the $p$-values of HISEI score and gender coefficients all lie above 0.05, so they are not statistically significant.

In classroom 1 we get a positive correlation between HISEI and literacy scores with $p$-value $<0.01$.


# Reproducibility Footer

```{r session-info}
cat("Seed used:", 161, "\n")
cat("Permutations used:", n_perm, "\n")
sessionInfo()
```
