---
title: "Assignment 2 — Task 3: SAOMs"
format: html
execute:
  cache: true
---

# Task 3: Estimation and Interpretation of SAOMs

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(RSiena)
library(sna)
library(network)

set.seed(161)

data_path <- "Glasgow"
```

#### Load Data

```{r load-data}
# Friendship adjacency matrices (no header, no row names)
f1 <- as.matrix(read.csv(file.path(data_path, "f1.csv"), header = FALSE))
f2 <- as.matrix(read.csv(file.path(data_path, "f2.csv"), header = FALSE))
f3 <- as.matrix(read.csv(file.path(data_path, "f3.csv"), header = FALSE))
dimnames(f1) <- dimnames(f2) <- dimnames(f3) <- NULL

# Demographic info: gender (1=boy, 2=girl), age
demographic <- read.csv(file.path(data_path, "demographic.csv"))
gender <- demographic$gender
age    <- demographic$age

# Alcohol consumption — three waves, ordinal 1..5; codebook does not include 0
alcohol <- as.matrix(read.csv(file.path(data_path, "alcohol.csv")))
dimnames(alcohol) <- NULL
alcohol[alcohol == 0] <- NA

# Log-distance dyadic covariate
logdistance <- as.matrix(read.csv(file.path(data_path, "logdistance.csv"), header = FALSE))
dimnames(logdistance) <- NULL

n <- nrow(f1)
stopifnot(nrow(f2) == n, nrow(f3) == n,
          length(gender) == n, nrow(alcohol) == n,
          nrow(logdistance) == n)
```

#### Summary Table

```{r summary-table}
summary_table <- data.frame(
  metric = c(
    "Pupils (n)",
    "Ties Wave 1", "Ties Wave 2", "Ties Wave 3",
    "Density Wave 1", "Density Wave 2", "Density Wave 3",
    "Boys", "Girls"
  ),
  value = c(
    n,
    sum(f1), sum(f2), sum(f3),
    round(mean(f1), 4), round(mean(f2), 4), round(mean(f3), 4),
    sum(gender == 1), sum(gender == 2)
  )
)
knitr::kable(summary_table, caption = "Glasgow Data Summary")
```

#### Geodesic Distribution Helper

```{r gof-helper}
# Auxiliary statistic for sienaGOF — not exported by RSiena
GeodesicDistribution <- function(i, data, sims, period, groupName, varName,
                                 levls = c(1:5, Inf), cumulative = TRUE, ...) {
  x <- networkExtraction(i, data, sims, period, groupName, varName)
  a <- sna::geodist(symmetrize(x))$gdist
  if (cumulative) {
    gdi <- sapply(levls, function(i) sum(a <= i))
  } else {
    gdi <- sapply(levls, function(i) sum(a == i))
  }
  names(gdi) <- as.character(levls)
  gdi
}
```

## Task 3.1 — Friendship Network as the Only Dependent Variable

### Task 3.1.1 — Jaccard Index
*Compute the Jaccard index to evaluate whether the data contain enough information to investigate the evolution of the friendship network.*

```{r task-3-1-1-jaccard}
jaccard <- function(A, B) {
  A <- ifelse(is.na(A), 0, A); B <- ifelse(is.na(B), 0, B)
  n11 <- sum(A == 1 & B == 1)
  n10 <- sum(A == 1 & B == 0)
  n01 <- sum(A == 0 & B == 1)
  n11 / (n11 + n10 + n01)
}

jac_table <- data.frame(
  Pair    = c("Wave 1 -> Wave 2", "Wave 2 -> Wave 3", "Wave 1 -> Wave 3"),
  Jaccard = c(jaccard(f1, f2), jaccard(f2, f3), jaccard(f1, f3))
)
knitr::kable(jac_table, digits = 4, caption = "Jaccard Index Between Waves")
```

The two consecutive-wave Jaccards (0.39 for 1->2 and 0.34 for 2->3) sit above the 0.30 threshold Snijders recommends for SAOM estimation, so the network changes gradually enough between observations for the SAOM "small steps" assumption to hold.

The Wave 1 to Wave 3 value (0.23) is lower because two years separate those snapshots; this is expected and not a problem, since estimation only uses the consecutive transitions. Overall the data have enough stability for the model to be identifiable and enough change for the dynamics to be worth modelling.

### Task 3.1.2 — Model Specification
*Specify a reasonable model to test:*\
*i. Students tend to be friends with popular pupils*\
*ii. Students tend to be friends with pupils with similar alcohol consumption*\
*iii. Students tend to be friends with students who live nearby (living nearby)*

```{r task-3-1-2-spec}
friendship <- sienaDependent(array(c(f1, f2, f3), dim = c(n, n, 3)))

genderCov  <- coCovar(as.numeric(gender))
ageCov     <- coCovar(as.numeric(age))
alcoholCov <- coCovar(as.numeric(alcohol[, 1]))   # Wave-1 alcohol as exogenous
logdistCov <- coDyadCovar(logdistance)

mydata1 <- sienaDataCreate(friendship, genderCov, ageCov, alcoholCov, logdistCov)

myeff1 <- getEffects(mydata1)

# Endogenous structural controls (in addition to default density + reciprocity)
myeff1 <- includeEffects(myeff1, transTrip)        # transitivity
myeff1 <- includeEffects(myeff1, cycle3)           # 3-cycles (anti-hierarchy)
# Hypothesis i — popularity
myeff1 <- includeEffects(myeff1, inPopSqrt)
# Hypothesis ii — alcohol homophily
myeff1 <- includeEffects(myeff1, simX, interaction1 = "alcoholCov")
# Hypothesis iii — living nearby (log-distance)
myeff1 <- includeEffects(myeff1, X,    interaction1 = "logdistCov")
# Exogenous gender controls
myeff1 <- includeEffects(myeff1, egoX, altX, sameX, interaction1 = "genderCov")
```

**Endogenous controls.** Reciprocity is already in the default. We add `transTrip` for transitivity and `cycle3` so transitivity is not confounded with generalised exchange.

**H-i (popularity).** We use `inPopSqrt` rather than `inPop`; the square-root form is less collinear with density and is the standard choice in the SAOM literature.

**H-iii (living nearby).** Log-distance enters as a dyadic covariate via the `X` effect. A *negative* coefficient on log-distance means closer pupils are more likely to be friends (positive "living nearby" effect).

### Task 3.1.3 — Estimation and Convergence

```{r task-3-1-3-estimate}
algo1 <- sienaAlgorithmCreate(projname = "saom_t31", seed = 161, n3 = 3000)

if (file.exists("model1.RData")) {
  load("model1.RData")
} else {
  model1 <- siena07(algo1, data = mydata1, effects = myeff1,
                    returnDeps = TRUE, batch = TRUE, verbose = FALSE)
  if (model1$tconv.max > 0.25 || max(abs(model1$tconv)) > 0.10) {
    model1 <- siena07(algo1, data = mydata1, effects = myeff1,
                      prevAns = model1, returnDeps = TRUE,
                      batch = TRUE, verbose = FALSE)
  }
  save(model1, file = "model1.RData")
}
```

```{r task-3-1-3-conv}
m1_conv <- data.frame(
  Statistic = c("Overall maximum convergence ratio (tconv.max)",
                "Maximum |t-ratio| per parameter"),
  Value     = c(round(model1$tconv.max, 3),
                round(max(abs(model1$tconv)), 3)),
  Threshold = c("< 0.25", "< 0.10")
)
knitr::kable(m1_conv, caption = "Model 1 — Convergence Diagnostics")
```

Both diagnostics are inside the RSiena thresholds (overall < 0.25, per-parameter < 0.10): `tconv.max = ` `r round(model1$tconv.max, 3)`, max |t| = `r round(max(abs(model1$tconv)), 3)`. The model converged on the first estimation pass; no continuation round was needed.

### Task 3.1.4 — Goodness of Fit
*Evaluate the goodness of fit of the model according to the auxiliary statistics discussed in the tutorial.*

```{r task-3-1-4-gof}
if (file.exists("gof1.RData")) {
  load("gof1.RData")
} else {
  gof1_indeg  <- sienaGOF(model1, IndegreeDistribution,  varName = "friendship", cumulative = FALSE)
  gof1_outdeg <- sienaGOF(model1, OutdegreeDistribution, varName = "friendship", cumulative = FALSE)
  gof1_geo    <- sienaGOF(model1, GeodesicDistribution,  varName = "friendship", cumulative = FALSE)
  gof1_tri    <- sienaGOF(model1, TriadCensus,           varName = "friendship")
  save(gof1_indeg, gof1_outdeg, gof1_geo, gof1_tri, file = "gof1.RData")
}
```

```{r task-3-1-4-gof-table}
gof1_p <- data.frame(
  Statistic = c("Indegree distribution", "Outdegree distribution",
                "Geodesic distribution", "Triad census"),
  `p-value` = c(round(gof1_indeg$Joint$p, 3),
                round(gof1_outdeg$Joint$p, 3),
                round(gof1_geo$Joint$p, 3),
                round(gof1_tri$Joint$p, 3)),
  check.names = FALSE
)
knitr::kable(gof1_p, caption = "Model 1 — sienaGOF p-values (Mahalanobis)")
```

```{r task-3-1-4-gof-plots, fig.width=10, fig.height=8}
par(mfrow = c(2, 2))
plot(gof1_indeg,  main = "GoF: Indegree Distribution")
plot(gof1_outdeg, main = "GoF: Outdegree Distribution")
plot(gof1_geo,    main = "GoF: Geodesic Distances")
plot(gof1_tri,    main = "GoF: Triad Census")
```

Reading the table: a p-value above 0.05 means the simulated and observed distributions are not distinguishable (i.e. the model fits that statistic).

- **Indegree** fits well (p = `r round(gof1_indeg$Joint$p, 3)`). This is consistent with `inPopSqrt` directly constraining the indegree shape.
- **Outdegree, geodesics, triads**: all p ≈ 0.

The likely cause is the questionnaire's six-friend cap. The empirical outdegree distribution is right-truncated at 6, but the linear `outdegree (density)` term cannot reproduce that truncation, so the simulated networks have too much spread on outdegree. Geodesics and the triad census inherit the misfit. Task 3.3 proposes candidate fixes (`outAct`/`outActSqrt`, `gwespFF`/`transTies`).

### Task 3.1.5 — Hypothesis Testing
*Are hypotheses i.–iii. supported by the data?*

```{r task-3-1-5-hyp}
m1_results <- data.frame(
  Effect   = model1$effects$effectName,
  Estimate = round(model1$theta, 4),
  SE       = round(sqrt(diag(model1$covtheta)), 4),
  t_value  = round(model1$theta / sqrt(diag(model1$covtheta)), 3)
)
knitr::kable(m1_results, caption = "Model 1 — Estimates")
```

**H-i — popularity. Not supported.**
`inPopSqrt` is *negative* and significant ($\hat\beta =$ `r round(model1$theta[5], 3)`, $t =$ `r round(model1$theta[5]/sqrt(diag(model1$covtheta))[5], 2)`). Once reciprocity and transitivity are in the model they absorb most of the "rich-get-richer" mass, and the residual popularity signal is slightly negative — popular pupils are not more likely to attract further ties.

**H-ii — alcohol homophily. Supported.**
`alcoholCov similarity` is positive and significant ($\hat\beta =$ `r round(model1$theta[10], 3)`, $t =$ `r round(model1$theta[10]/sqrt(diag(model1$covtheta))[10], 2)`). Pupils prefer to nominate peers with similar Wave-1 alcohol levels.

**H-iii — living nearby. Supported.**
Log-distance has a negative significant effect ($\hat\beta =$ `r round(model1$theta[6], 3)`, $t =$ `r round(model1$theta[6]/sqrt(diag(model1$covtheta))[6], 2)`). Greater distance lowers the probability of a friendship tie.

**Other notable effects.**

- Reciprocity ($\hat\beta =$ `r round(model1$theta[2], 2)`) and transitive triplets ($\hat\beta =$ `r round(model1$theta[3], 2)`) are strongly positive — both expected.
- 3-cycles is negative ($\hat\beta =$ `r round(model1$theta[4], 2)`), pointing to a hierarchical rather than generalised-exchange structure.
- Same-gender homophily is large ($\hat\beta =$ `r round(model1$theta[9], 2)`); the gender ego and alter main effects are not individually significant.

## Task 3.2 — Co-evolution of Friendship and Alcohol

### Task 3.2.1 — Model Specification (Selection + Influence)
*Use the network specification from (1) for the selection part. Specify the influence part to test:*\
*iv. Popular students tend to increase or maintain their level of alcohol consumption*\
*v. Students tend to adjust their alcohol consumption to that of their friends*

```{r task-3-2-1-spec}
friendship2 <- sienaDependent(array(c(f1, f2, f3), dim = c(n, n, 3)))
drinkingbeh <- sienaDependent(alcohol, type = "behavior")

genderCov2  <- coCovar(as.numeric(gender))
ageCov2     <- coCovar(as.numeric(age))
logdistCov2 <- coDyadCovar(logdistance)

mydata2 <- sienaDataCreate(friendship2, drinkingbeh, genderCov2, ageCov2, logdistCov2)

myeff2 <- getEffects(mydata2)

# --- Selection part — same as Model 1, but alcohol is now the behavioural DV ---
myeff2 <- includeEffects(myeff2, transTrip, cycle3)
myeff2 <- includeEffects(myeff2, inPopSqrt)
myeff2 <- includeEffects(myeff2, egoX, altX, sameX, interaction1 = "genderCov2")
myeff2 <- includeEffects(myeff2, X,    interaction1 = "logdistCov2")
myeff2 <- includeEffects(myeff2, simX, interaction1 = "drinkingbeh")

# --- Influence part ---
# Hypothesis iv — popular students change their drinking
myeff2 <- includeEffects(myeff2, indeg,
                         name = "drinkingbeh", interaction1 = "friendship2")
# Hypothesis v — students adjust to the average of their friends
myeff2 <- includeEffects(myeff2, avSim,
                         name = "drinkingbeh", interaction1 = "friendship2")
# Behavioural control — gender main effect on drinking
myeff2 <- includeEffects(myeff2, effFrom,
                         name = "drinkingbeh", interaction1 = "genderCov2")
```

The selection block mirrors Model 1, with one change: alcohol homophily now uses the *dynamic* behavioural variable (`simX` on `drinkingbeh`) instead of the Wave-1 snapshot.

The behaviour function already includes the linear and quadratic shape by default. On top of those:

- `indeg` for H-iv (popularity → drinking)
- `avSim` for H-v (friends' drinking → ego's drinking)
- `effFrom(genderCov2)` as a control on the behaviour

### Task 3.2.2 — Estimation and Convergence

```{r task-3-2-2-estimate}
algo2 <- sienaAlgorithmCreate(projname = "saom_t32", seed = 161, n3 = 3000)

if (file.exists("model2.RData")) {
  load("model2.RData")
} else {
  model2 <- siena07(algo2, data = mydata2, effects = myeff2,
                    returnDeps = TRUE, batch = TRUE, verbose = FALSE)
  tries <- 0
  while ((model2$tconv.max > 0.25 || max(abs(model2$tconv)) > 0.10) && tries < 3) {
    tries <- tries + 1
    model2 <- siena07(algo2, data = mydata2, effects = myeff2,
                      prevAns = model2, returnDeps = TRUE,
                      batch = TRUE, verbose = FALSE)
  }
  save(model2, file = "model2.RData")
}
```

```{r task-3-2-2-conv}
m2_conv <- data.frame(
  Statistic = c("Overall maximum convergence ratio (tconv.max)",
                "Maximum |t-ratio| per parameter"),
  Value     = c(round(model2$tconv.max, 3),
                round(max(abs(model2$tconv)), 3)),
  Threshold = c("< 0.25", "< 0.10")
)
knitr::kable(m2_conv, caption = "Model 2 — Convergence Diagnostics")
```

`tconv.max =` `r round(model2$tconv.max, 3)`, max |t| per parameter `=` `r round(max(abs(model2$tconv)), 3)`. Both clear the 0.25 / 0.10 thresholds, so the joint network–behaviour model converged.

`tconv.max` is higher than in Model 1 (`r round(model1$tconv.max, 3)`), but that is expected: Model 2 has 19 parameters instead of 10 and adds a behavioural dependent variable, which makes the moment-matching problem stiffer.

### Task 3.2.3 — Goodness of Fit

```{r task-3-2-3-gof}
if (file.exists("gof2.RData")) {
  load("gof2.RData")
} else {
  gof2_indeg  <- sienaGOF(model2, IndegreeDistribution,  varName = "friendship2", cumulative = FALSE)
  gof2_outdeg <- sienaGOF(model2, OutdegreeDistribution, varName = "friendship2", cumulative = FALSE)
  gof2_geo    <- sienaGOF(model2, GeodesicDistribution,  varName = "friendship2", cumulative = FALSE)
  gof2_beh    <- sienaGOF(model2, BehaviorDistribution,  varName = "drinkingbeh")
  save(gof2_indeg, gof2_outdeg, gof2_geo, gof2_beh, file = "gof2.RData")
}
```

```{r task-3-2-3-gof-table}
gof2_p <- data.frame(
  Statistic = c("Indegree distribution", "Outdegree distribution",
                "Geodesic distribution", "Behaviour distribution"),
  `p-value` = c(round(gof2_indeg$Joint$p, 3),
                round(gof2_outdeg$Joint$p, 3),
                round(gof2_geo$Joint$p, 3),
                round(gof2_beh$Joint$p, 3)),
  check.names = FALSE
)
knitr::kable(gof2_p, caption = "Model 2 — sienaGOF p-values (Mahalanobis)")
```

```{r task-3-2-3-gof-plots, fig.width=10, fig.height=8}
par(mfrow = c(2, 2))
plot(gof2_indeg,  main = "GoF: Indegree Distribution")
plot(gof2_outdeg, main = "GoF: Outdegree Distribution")
plot(gof2_geo,    main = "GoF: Geodesic Distances")
plot(gof2_beh,    main = "GoF: Alcohol Distribution")
```

Same pattern as Model 1 on the network side, with one new fit to evaluate (the behaviour distribution):

- **Indegree**: fits (p = `r round(gof2_indeg$Joint$p, 3)`).
- **Behaviour distribution**: fits well (p = `r round(gof2_beh$Joint$p, 3)`) — the alcohol side of the model is adequate.
- **Outdegree and geodesics**: still p ≈ 0, same six-nomination-cap issue. Task 3.3 covers candidate fixes.

### Task 3.2.4 — Hypothesis Testing
*Are hypotheses iv.–v. supported? How do conclusions about i.–iii. differ from (1)?*

```{r task-3-2-4-hyp}
m2_results <- data.frame(
  Effect   = model2$effects$effectName,
  Estimate = round(model2$theta, 4),
  SE       = round(sqrt(diag(model2$covtheta)), 4),
  t_value  = round(model2$theta / sqrt(diag(model2$covtheta)), 3)
)
knitr::kable(m2_results, caption = "Model 2 — Estimates")
```

**H-iv — popular students drink more. Not supported.**
`drinkingbeh indegree` is small and not significant ($\hat\beta =$ `r round(model2$theta[18], 3)`, $t =$ `r round(model2$theta[18]/sqrt(diag(model2$covtheta))[18], 2)`). Receiving many nominations does not push drinking up, at least once `avSim` (the influence channel) is in the model.

**H-v — assimilation to friends' drinking. Supported.**
`drinkingbeh average similarity` is large and significant ($\hat\beta =$ `r round(model2$theta[17], 2)`, $t =$ `r round(model2$theta[17]/sqrt(diag(model2$covtheta))[17], 2)`). Pupils whose alcohol level differs from their friends' average move toward it over time.

**Revisiting H-i to H-iii in the joint model.**
The three selection-side coefficients are close to Model 1:

- **H-i (popularity):** still negative and significant ($\hat\beta =$ `r round(model2$theta[7], 3)`, $t =$ `r round(model2$theta[7]/sqrt(diag(model2$covtheta))[7], 2)`) — still **not supported**.
- **H-iii (living nearby):** still negative and significant on log-distance ($\hat\beta =$ `r round(model2$theta[8], 3)`, $t =$ `r round(model2$theta[8]/sqrt(diag(model2$covtheta))[8], 2)`) — still **supported**.
- **H-ii (alcohol homophily):** now uses the evolving behaviour and is *stronger* than in Model 1 ($\hat\beta =$ `r round(model2$theta[12], 3)` vs `r round(model1$theta[10], 3)`; $t =$ `r round(model2$theta[12]/sqrt(diag(model2$covtheta))[12], 2)`) — still **supported**.

The bigger homophily coefficient under the dynamic DV is plausible: in Model 1 the Wave-1 snapshot is a noisy proxy for "alcohol level at the time the tie formed", which attenuates the estimate. Using the dynamic DV removes that attenuation.

### Task 3.2.5 — Selection vs. Influence
*Given the model in (2.2), do we have evidence for selection only, influence only, or both?*

Both mechanisms show up in the model:

- **Selection** (`simX(drinkingbeh)`): $\hat\beta =$ `r round(model2$theta[12], 2)`, $t =$ `r round(model2$theta[12]/sqrt(diag(model2$covtheta))[12], 2)`.
- **Influence** (`avSim` on drinking): $\hat\beta =$ `r round(model2$theta[17], 2)`, $t =$ `r round(model2$theta[17]/sqrt(diag(model2$covtheta))[17], 2)`.

So we find evidence for **both selection and influence**. Pupils both pick friends whose drinking is similar to theirs (selection) *and* shift their own drinking toward their friends' average over time (influence).

This is the case where SAOMs are useful: cross-sectional homophily on alcohol is consistent with either mechanism alone, so without the joint dynamic model we could not separate them. The two effects are identifiable jointly because they act on different parts of the data — selection on ties given attributes, influence on attribute changes given ties.

## Task 3.3 — Model Improvement
*Discuss new effects that could improve the fit on geodesic distances and degree distributions.*

The fit problems sit on three statistics: outdegree, geodesics, and (in Model 1) the triad census. They share a root cause — the questionnaire's six-friend cap, which a linear `outdegree (density)` term cannot reproduce. Once the outdegree distribution is wrong, geodesics and the triad census inherit the misfit.

**Outdegree shape.** Add `outActSqrt` (or `outAct`, or a `gwodegree`-style term). With a negative coefficient these saturate the propensity to send another tie, mimicking the cap. An `outdegree truncated at 6` term would do something similar by hard-coding the truncation point.

**Indegree shape.** `inActSqrt` lets the rate of receiving further nominations bend non-linearly with current indegree, which is more flexible than the strictly linear assumption built into `inPopSqrt`.

**Triadic and geodesic fit.** Replace (or supplement) `transTrip` with `gwespFF` or `transTies`. `transTrip` over-rewards ties that close *many* transitive paths simultaneously; `gwespFF` flattens that and tends to match the observed clustering distribution more reliably.

**Other directions.**

- `balance` or `transRecTrip` to capture reciprocity-transitivity interactions, which are part of the triad census the current model misses.
- An `absDiffX` on age (a dyadic age-difference covariate) — friendships within cohorts often cluster, and tighter clusters tend to improve the geodesic fit.

## Task 3.4 — Two Additional Hypotheses
*Two further testable hypotheses on friendship/alcohol dynamics, and how to operationalise them.*

**Hypothesis A — Drinking and smoking co-evolve.**
Adolescent alcohol and tobacco use are often correlated. Beyond the network-mediated channels, the two behaviours may push each other directly.

*Operationalisation.* Add smoking as a second behavioural DV (`sienaDependent(smoke, type = "behavior")`) and include the cross-behaviour effects `effFrom(drinkingbeh, smoke)` and `effFrom(smoke, drinkingbeh)`. A positive cross-effect would mean drinking pushes up smoking and vice versa, on top of the network-driven influence and selection. Requires per-wave smoking data; otherwise a standard SAOM extension.

**Hypothesis B — Heavy drinkers have less stable friendships.**
Alcohol use may shape not only *who* a pupil is friends with, but also how stable those friendships are: heavier drinkers may churn through ties faster.

*Operationalisation.* Either let the network rate depend on alcohol via `effFrom(rate, drinkingbeh)`, or split `egoX(drinkingbeh)` into its `creation` and `endowment` (tie-maintenance) variants. A *negative endowment* effect (or *positive rate* effect of drinking) would mean heavy drinkers are more likely to drop existing ties, separating tie *formation* from tie *persistence*.

## Reproducibility Footer

```{r session-info}
cat("Seed used: 161\n")
sessionInfo()
```
