---
title: "Assignment 1"
format: html
execute:
  cache: true
---

# Task 1: Tie Dependence and CUGs

## Setup

```{r setup}
# Given by task statement
set.seed(161)
n_perm <- 5000

m <- 134
m <- 42
a <- 50
n <- 139

# Data path from this folder to the project Data folder
data_path <- "Lintner_Data/10_W1.csv"
```

#### Load Network and Remove Affected Data

```{r load-network}
# 1) Load network with blanks as NA
raw_net <- read.csv2(
  "../Data/10_W1.csv",   # adjust path if needed
  header = TRUE,
  row.names = 1,
  check.names = FALSE,
  na.strings = c("", "NA")
)

# Keep original node labels and normalize possible "X" prefixes in numeric IDs
row_ids <- rownames(raw_net)
col_ids <- colnames(raw_net)

# Convert to numeric matrix and align to common IDs on both axes
A <- as.matrix(raw_net)
storage.mode(A) <- "numeric"
rownames(A) <- row_ids
colnames(A) <- col_ids

common_ids <- intersect(row_ids, col_ids)
A <- A[common_ids, common_ids, drop = FALSE]

# 2) Identify nodes affected by missing tie data
nodes_drop <- rownames(A)[rowSums(is.na(A)) > 0]

# 3) Remove affected nodes from network (row + column)
A_clean <- A[!(rownames(A) %in% nodes_drop), !(colnames(A) %in% nodes_drop), drop = FALSE]

# 4) Final checks
stopifnot(nrow(A_clean) == ncol(A_clean))
stopifnot(all(A_clean %in% c(0, 1)))
```

#### Summary Table

```{r summary-table}
summary_table <- data.frame(
  metric = c("Original nodes", "Dropped nodes", "Cleaned nodes", "Remaining NA", "Ties in cleaned network"),
  value = c(
    nrow(A),
    length(nodes_drop),
    nrow(A_clean),
    sum(is.na(A_clean)),
    sum(A_clean)
  )
)

nodes_drop_table <- data.frame(
  dropped_node = if (length(nodes_drop) == 0) NA_character_ else nodes_drop,
  stringsAsFactors = FALSE
)

knitr::kable(summary_table, caption = "Cleaning Summary")
knitr::kable(nodes_drop_table, caption = "Dropped Node IDs")
```

#### Cleaned 10_W1 CSV 

```{r save-clean-csv}
# Save cleaned adjacency matrix as a CSV next to this qmd
clean_csv_path <- "10_W1_clean.csv"
write.csv2(
  data.frame(node = rownames(A_clean), A_clean, check.names = FALSE),
  file = clean_csv_path,
  row.names = FALSE,
  quote = FALSE
)
```

## Task 1.1
*Compute probability $p$ of observing a tie if you choose one possible dyad at random*

```{r task-1-1-tie-prob}
nodes = nrow(raw_net) # We work with the raw input (not the cleaned A_clean)
ties  = sum(raw_net, na.rm = TRUE)

# P[tie] = |ties| / |total number of possible ties|
n_dyads = (nodes) * (nodes - 1) / 2
p = ties / n_dyads
cat("Probability[Observing a Tie] = ", p, "\n")
```

## Task 1.2
*Compute the probabilities $p_M, p_A, p_N$ of observing a mutual, asymmetric and null dyad*

```{r task-1-2-dyad-probs}
# Convert to numeric matrix, treating NA as missing
A_raw <- as.matrix(raw_net)
storage.mode(A_raw) <- "numeric"

# Get upper triangle indices (each dyad counted once, i < j)
upper <- upper.tri(A_raw)
ij <- A_raw[upper]          # tie from i -> j
ji <- t(A_raw)[upper]       # tie from j -> i

n_mutual     <- sum(ij == 1 & ji == 1, na.rm = TRUE)
n_asymmetric <- sum(xor(ij == 1, ji == 1), na.rm = TRUE)
n_null       <- sum(ij == 0 & ji == 0, na.rm = TRUE)
n_nan        <- n_dyads - (n_mutual + n_asymmetric + n_null)

p_M <- n_mutual     / n_dyads
p_A <- n_asymmetric / n_dyads
p_N <- n_null       / n_dyads
p_Na <- n_nan / n_dyads

knitr::kable(
  data.frame(
    Dyad        = c("Mutual", "Asymmetric", "Null", "Rest (NA)", "Total"),
    Count       = c(n_mutual, n_asymmetric, n_null, n_nan, n_dyads),
    Probability = c(p_M, p_A, p_N, p_Na, p_M + p_A + p_N + p_Na)
  ),
  digits = 4,
  caption = "Dyad Census"
)
```

## Task 1.3
*What would be the value of the probabilities in (2) if we assume tie independence and that the probability of observing a tie takes the value $p$ computed in (1)?*

```{r task-1-3-tie-prob-uniform}
# T : Tie
# N : No Tie
# 
# p[T] = p
# p[N] = 1 - p
# 
#      . 
#     / \ 
#   T     N
#  /\     /\   
# t  N   T  N

p_uni_M <- p * p
p_uni_A <- 2 * p * (1 - p)
p_uni_N <- (1 - p) * (1 - p)

knitr::kable(
  data.frame(
    Dyad        = c("Mutual", "Asymmetric", "Null"),
    Probability = c(p_uni_M, p_uni_A, p_uni_N)
  ),
  digits = 4,
  caption = "Dyad Census"
)
```

## Task 1.4
*Given the values obtained in (2) and (3), would it be reasonable to assume tie independence when performing further analysis on the observed network? Justify*

No definitely not. The individual probabilities of mutual, asymmetric and null diads do not even match closely.

## Task 1.5
*Consider a different model that assumes all nodes have an outdegree of nine (d=9).*
*Compute the probability of a mutual tie assuming tie independence. Compare this probability with the probabilities computed in (2) and (3). Would it be reasonable to assume this model?*

```{r task-1-5-different-model}

```


## Reproducibility Footer

```{r session-info}
cat("Seed used:", 161, "\n")
cat("Permutations used:", n_perm, "\n")
sessionInfo()
```

