# Idea

The Network Autocorrelation Model assumes the data is distributed like:
$$y = X\beta + \rho W y + \epsilon,$$
where in our case $y$ is the ending literacy score, $X$ is the other attributes, $W$ is the wave 2 adjacency matrix and $\epsilon$ is distributed normally.

So we just need to statistically estimate $\beta, \rho$.

# Importing and cleaning Data

We need to clean the data by removing all students with missing data. That can be missing friendship data, missing HISEI, missing gender, missing literacy, ...

The important code-block for this cleaning is the following. We get all student ids and intersect them to get only the ones with complete data.

```{r}
#| eval: false
ok_ids1 <- rownames(w2net)[complete.cases(w2net)]
ok_ids2 <- rownames(attr)[complete.cases(attr)]

ok_ids <- intersect(ok_ids1, ok_ids2)
```

The remaining code of `import_and_clean.R` will be uploaded along the pdf.

```{r}
#| message: false
source("import_and_clean.R")
w2netClean<-getNet("10")
attrClean<-getAttr("10")
```

# Subtask 1

Subtask 1 requires just Network Autocorrelation, without exogenous variables, so the model simplifies to
$$y = \rho W y + \epsilon.$$

So we fit a NAM model:
```{r}
model <- lnam(y=attrClean$literacy_end, x=NULL, W1=w2netClean)
print(model)
```

# Subtask 2

Subtask 2 adds HISEI and gender, giving the model described in **Idea**.

Gender is a simple binary variable, encoding it as $0$ for male and $1$ for female should be sufficient:

```{r}
attrClean["gender"] <- lapply(attrClean["gender"], \(x) x=="female")
```

Now we can basically reuse the code from **Subtask 1**, just adding gender and HISEI in the $X$ matrix.

```{r}
X <- model.matrix(literacy_end~ gender + HISEI, data = attrClean)
model <- lnam(attrClean$literacy_end, X, w2netClean)
print(model)
```

# Subtask 3

The outcome actually suggests that the autocorrelation term is negatively correlated.
<!-- But the value of the coefficient is very small, so it is not really a relevant result. -->

# Subtask 4

TODO (discuss with others)

# Subtask 5

```{r}
for (classroom in c("01", "02", "03", "04", "05", "09", "10", "12")) {
    w2netClean<-getNet(classroom)
    attrClean<-getAttr(classroom)

    attrClean["gender"] <- lapply(attrClean["gender"], \(x) x=="female")    

    X <- model.matrix(literacy_end~ gender + HISEI, data = attrClean)
    model <- lnam(attrClean$literacy_end, X, w2netClean)
    print(model)
}
```

# Subtask 6

