# Idea

The Network Autocorrelation Model assumes the data is distributed like:
$$y = X\beta + \rho W y + \epsilon,$$
where in our case $y$ is the ending literacy score, $X$ is the other attributes, $W$ is the wave 2 adjacency matrix and $\epsilon$ is distributed normally.

So we just need to statistically estimate $\beta, \rho$.

# Importing and cleaning Data

```{r}
#| message: false
source("import_and_clean.R")
w2netClean<-getNet("10")
attrClean<-getAttr("10")
```

# Subtask 1

Subtask 1 requires just Network Autocorrelation, without exogenous variables, so the model simplifies to
$$y = \rho W y + \epsilon.$$

We first add the Autocorrelation to the attributes dataframe:
```{r}
autocorrelation <- w2netClean %*% as.numeric(as.vector(attrClean[,"literacy_end"]))
attrClean["autocorrelation"] <- autocorrelation
```

And then fit a model:
```{r}
model <- lm(literacy_end ~ autocorrelation, attrClean)
print(model)
plot(attrClean$autocorrelation, attrClean$literacy_end)
abline(model)
```

# Subtask 2

Subtask 2 adds HISEI and gender, giving the model described in **Idea**.

Gender is a simple binary variable, encoding it as $0$ for male and $1$ for female should be sufficient:

```{r}
attrClean["gender"] <- lapply(attrClean["gender"], \(x) x=="female")
print(attrClean)
```

Now we can basically reuse the code from **Subtask 1**, just adding gender and HISEI.

```{r}
model <- lm(literacy_end ~ autocorrelation + HISEI + gender, attrClean)
print(model)
```

# Subtask 3

The outcome actually suggests that the autocorrelation term is negatively correlated.
But the value of the coefficient is very small, so it is not really a relevant result.

# Subtask 4

TODO (discuss with others)

# Subtask 5

```{r}
for (classroom in c("01", "02", "03", "04", "05", "09", "10", "12")) {
    w2netClean<-getNet(classroom)
    attrClean<-getAttr(classroom)

    autocorrelation <- w2netClean %*% as.numeric(as.vector(attrClean[,"literacy_end"]))
    attrClean["autocorrelation"] <- autocorrelation

    attrClean["gender"] <- lapply(attrClean["gender"], \(x) x=="female")    

    model <- lm(literacy_end ~ autocorrelation + HISEI + gender, attrClean)
    print(model)
}
```

# Subtask 6

WTF??


TODO: FIX SEED!!