# Idea

The Network Autocorrelation Model assumes the data is distributed like:
$$y = X\beta + \rho W y + \epsilon,$$
where in our case $y$ is the ending literacy score, $X$ is the other attributes, $W$ is the wave 2 adjacency matrix and $\epsilon$ is distributed normally.

So we just need to statistically estimate $\beta, \rho$.

# Importing and cleaning Data

We need to clean the data by removing all students with missing data. That can be missing friendship data, missing HISEI, missing gender, missing literacy, ...

The important code-block for this cleaning is the following. We get all student ids and intersect them to get only the ones with complete data.

```{r}
#| eval: false
ok_ids1 <- rownames(w2net)[complete.cases(w2net)]
ok_ids2 <- rownames(attr)[complete.cases(attr)]

ok_ids <- intersect(ok_ids1, ok_ids2)
```

The remaining code of `import_and_clean.R` will be uploaded along the pdf.

```{r}
#| message: false
source("import_and_clean.R")
w2netClean<-getNet("10")
attrClean<-getAttr("10")
```

# Subtask 1

Subtask 1 requires just Network Autocorrelation, without exogenous variables, so the model simplifies to
$$y = \mathbf{1} \beta + \rho W y + \epsilon.$$
for some constants $\beta, \rho$ to be estimated.

So we fit a NAM model:
```{r}
X <- model.matrix(literacy_end~1, data=attrClean)
model <- lnam(y=attrClean$literacy_end, x=X, W1=w2netClean)
print(summary(model))
```

# Subtask 2

Subtask 2 adds HISEI and gender, giving the model described in **Idea**.

Gender is a simple binary variable, encoding it as $0$ for male and $1$ for female should be sufficient:

```{r}
attrClean["gender"] <- lapply(attrClean["gender"], \(x) x=="female")
```

Now we can basically reuse the code from **Subtask 1**, just adding gender, HISEI and a constant factor for the intersect in the $X$ matrix.

```{r}
X <- model.matrix(literacy_end~ gender + HISEI, data = attrClean)
model <- lnam(attrClean$literacy_end, X, w2netClean)
print(summary(model))
```

# Subtask 3

For classroom 10, we get a negative autocorrelation term with $p$-value $0.028$. So it's significant. This doesn't support the positive autocorrelation hypothesis, it actually contradicts it. This suggests that students are likely connected with other students which have differing literacy scores. Further, being female and having a high HISEI score both correlate positively with the literacy_end score.

<!-- The outcome actually suggests that the autocorrelation is negative. Further, female students outperform male students, and there is a strong correlation between the HISEI score and the literacy_end score. -->

# Subtask 4

**TODO (discuss with others)**

QAP and NAM address different questions. The QAP task examines whether different variables have an effect on the probability of a tie existing, while NAM addresses the question wether different variables (including the network structure itself) have an effect on the ending literacy score.

<!-- Comparing these results with the results from QAP, we see that QAP also revealed that female students perform better, and that there is a correlation between HISEI and literacy_end scores. -->

# Subtask 5

Combining all the above code and running it in a loop over all classrooms, we get:

```{r}
for (classroom in c("01", "02", "03", "04", "05", "09", "10", "12")) {
    w2netClean<-getNet(classroom)
    attrClean<-getAttr(classroom)

    attrClean["gender"] <- lapply(attrClean["gender"], \(x) x=="female")    

    X <- model.matrix(literacy_end~ gender + HISEI, data = attrClean)
    model <- lnam(attrClean$literacy_end, X, w2netClean)
    print(summary(model))
}
```

# Subtask 6

In the results from subtask 5 most of the autocorrelation terms are negative, but all with very large $p$-value. So the result is not statistically significant. Only classroom 10 has evidence for such a negative autocorrelation. These results are not generalizable. So there is no evidence to support a positive autocorrelation hypothesis.
